Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-12 content/contribution/continuous-datalog.tex
Fixed definitions for Datalog, time. Reordered to move window definitions after TA-Datalog

%% fixme
%	In regards to Continuous Datalog, an ``instant'', a ``stream of facts'' and a ``sliding window'' over a stream are defined as follows:

@@

\[cons(\Pi) = \{ f \in \hat{H} | \Pi \models f \}\]
\end{definition}

\begin{definition}[Justification]
\labeldef{continuous datalog: justification}
A {\em justification} for the entailment $E$ of some Datalog program
${\langle R , F \rangle \models E}$ is a minimal set
${J \subseteq F}$ such that ${\langle R , J \rangle \models E}$.
\end{definition}

\begin{definition}[Instant]
\labeldef{continuous datalog: instant}
We assume a discrete, ordered time domain $\mathbb{T}$; an instant $i
\in \mathbb{T}$ is a point in time. It should be noted that time is a
continuous concept, with infinitely many divisions possible between
any two discrete instants.  However, the resolution with which such
divisions can be recognised will, in practise, have a lower bound set
by hardware limitations, or may be defined as part of some specific
implementation.  In these cases, anything that occurs between discrete
instants is considered to have occurred at the next recognised
instant.
\end{definition}

\begin{definition}[Fact Instance]
\labeldef{continuous datalog: fact instance}
A {\em fact instance} is a fact with some additional temporal
annotation; it is a tuple containing a fact $f$ and the timestamp
$t \in \mathbb{T}$ at which $f$ became true, written
$\langle f, t \rangle$. Two fact instances containing the same fact
and timestamp are considered identical.

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function ${strip : \hat{H} \times \mathbb{T} \rightarrow \hat{H}}$
which maps a fact instance to its underlying fact.
\end{definition}

@@

In \refdef{continuous datalog: datalog program}, we took a
Datalog program to consist of a set of rules and a set of facts.
Taken together with the set of rules $R$, the set of facts $F$ entails
another set of facts $E$ in which the existence of one fact in $E$ may
be justified by many distinct subsets of $F$.

In order to extend this
definition of Datalog programs to incorporate fact insatnces,
as in \refdef{continuous datalog: fact instance}, we define a
time-annotated Datalog program in \refdef{continuous datalog: ta datalog program}.

\begin{definition}[Time-Annotated Datalog Program]
\labeldef{continuous datalog: ta datalog program}
A time-annotated Datalog program $\Pi^T = \langle R, F^T\rangle$ is a tuple
which comprises a finite set of rules $R$ and a finite set of fact instances
$F^T$. Fact instances are extensional atomic formulae of first-order logic
annotated with a timestamp, as in \refdef{continuous datalog: fact instance},
and rules are of the same form as those of conventional Datalog programs.

As discussed in \refdef{continuous datalog: datalog program,continuous datalog: fact instance},
in the context of the Semantic Web we treat fact instances as time-annotated
3-tuples ${\langle s, p, o \rangle | t}$.
\end{definition}

When considering a time-annotated Datalog program, in which there may be
many distinct instances of a given fact, temporal information regarding when
instances of that fact became true would be lost if such a program were to entail
sets of plain facts. As such, where a fact $f$ would be entailed by some Datalog
program ${\Pi = \langle R, \{ strip(x) : x \in W^{S,x}_i \} \rangle}$,
for some window over some stream $S$, an instance ${f | t}$ of $f$ should be
entailed by the time-annotated Datalog program
${\Pi^T = \langle R, W^{S,x}_i \rangle \vdash f | t}$.

It follows that applying the ${strip()}$ function to all the fact instances in
both the set of fact instances $F^T$ in some ${\Pi^T = \langle R, F^T \rangle \vDash E^T}$
and the set of entailed fact instances $E^T$ maintains the relationship of entailment:
\begin{multline*}
	\left( \langle R, F^T \rangle \vDash E^T \right) \rightarrow \\
		\left( \langle R, \{ strip(x) : x \in F^T \} \rangle \vDash \{ strip(x) : x \in E^T \} \right)
\end{multline*}

It should be noted that \refdef{continuous datalog: justification} of entailment
justification is equally applicable to time-annotated Datalog programs,
simply replacing entailed and justifying facts with fact instances.
Furthermore, each justification for $f$ in some $F^T$ entails a unique
instance ${f | t}$, such that many instances of $f$ may occur in the set
$E^T$ of fact instances entailed by
${\Pi^T = \langle R, W^{S,x}_i \rangle}$, each with a single
justification.  As such, entailment justification for fact instances may be
defined formally by \refdef{continuous datalog: instance justification}.
\begin{definition}[Instance Entailment Justification]
\labeldef{continuous datalog: instance justification}
	A justification for some entailment ${f|t}$ of some time-annotated Datalog program
	${\Pi^T = \langle R , F^T \rangle \vdash f|t}$ is the single minimal set ${J^T \subseteq F^T}$
	such that ${\Pi^{T\prime} = \langle R , J^T \rangle \vdash f|t}$.
	The mapping of the fact instance ${f | t}$ to the minimal subset $J^T$
	within ${\Pi^T = \langle R, F^T \rangle}$ that justifies it is expressed
	as the function ${just : \mathbb{\Pi^T} , \mathbb{F^T} \rightarrow \{ \mathbb{F^T} \}}$.
\end{definition}

>>>>

@@

\[cons(\Pi) = \{ f \in \hat{H} | \Pi \models f \}\]

%For convenience, we will write $\Pi \models E$ to indicate that $E =
%cons(\Pi)$ is the set of all facts entailed by $\Pi$.

\begin{definition}[Justification]
\labeldef{continuous datalog: justification}

We define a {\em justification} of an entailment of a Datalog program
as a minimal subset of the program that is sufficient for the
entailment to hold. More precisely, for a Datalog program
$\Pi = \langle R, F \rangle$ and an entailment $E$ where
$\langle R, F \rangle \models E$, the program $\langle J_R, J_F \rangle$
with $J_R \subseteq R$ and $J_F \subseteq F$ is a justification for $E$
iff $\langle J_R, J_F \rangle \models E$ and, for all $J'_R \subset J_R$ or $J'_F \subset J_F$, $\langle J'_R, J'_F \rangle \not\models E$.

We can simplify this notion of justification by making the assumption
that the justification can be defined in terms of a minimal subset of
the ground facts of the program; the rules of the program are
therefore implicitly considered to be part of every justification. We
refer to this simplified justification as a {\em fact justification},
and define it as follows: given a Datalog program $\Pi = \langle R, F
\rangle$ and an entailment $E$ where $\langle R, F \rangle \models E$,
the set of facts $J \subseteq F$ is a fact justification for $E$ iff
$\langle R, J \rangle \models E$ and, for all $J' \subset J$, $\langle R, J' \rangle
\not\models E$.
\end{definition}

\begin{definition}[Time Domain]
\labeldef{continuous datalog: instant}

Let $\mathbb{T}$ be a discrete time domain with a total order under
$\leqslant$; an element of $\mathbb{T}$ is an {\em instant}.
$\mathbb{T}$ has a lower bound $0$, such that for all $i \in
\mathbb{T}$, $0 \leqslant i$. For convenience, we write $i+n$ where $i
\in \mathbb{T}$ to indicate the $n$th successor of instant $i$,
similarly $i-n$.

%It should be noted that time is a continuous concept, with infinitely
%many divisions possible between any two discrete instants.  However,
%the resolution with which such divisions can be recognised will, in
%practise, have a lower bound set by hardware limitations, or may be
%defined as part of some specific implementation.  In these cases,
%anything that occurs between discrete instants is considered to have
%occurred at the next recognised instant.
\end{definition}

\begin{definition}[Fact Instance]
 \labeldef{continuous datalog: fact instance}

A {\em fact instance} is a fact with some additional temporal
annotation; it is a tuple containing a fact $f$ and an instant $t \in
\mathbb{T}$, written $\langle f, t \rangle$. Two fact instances
containing the same fact and timestamp are considered identical; two
fact instances containing the same fact but different timestamps are
considered to be distinct.

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function ${strip : \hat{H} \times \mathbb{T} \rightarrow \hat{H}}$
which maps a fact instance to its underlying fact:

\[ strip(\langle f, t\rangle) = f \]

For convenience, we also write $strip(F^T) = F$, where $F^T$ is a set
of fact instances and $F = \{ strip(f) | f \in F^T \}$.

\begin{definition}[Time-Annotated Datalog]
\labeldef{continuous datalog: ta datalog program}

A time-annotated Datalog program $\Pi^T = \langle R, F^T\rangle$ is a
tuple which comprises a finite set of rules $R$ and a finite set of
fact instances $F^T \subseteq \hat{H} \times \mathbb{T}$. Fact
instances are as defined in \refdef{continuous datalog: fact
  instance}, and rules are of the same form as in conventional
Datalog programs.

When considering a time-annotated Datalog program, in which there may
be many distinct instances of a given fact, temporal information
regarding when instances of that fact became true would be lost if
such a program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts at all
instants). As before, a fact $f$ is entailed by a program $\Pi^T$ iff
each interpretation satisfying every clause in $\Pi^T$ also satisfies
$f$. The entailment of $\Pi^T$ is therefore:

\[ cons^T(\Pi^T) = \{ f \in \hat{H}^T | \Pi^T \models f \} \]

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the set of entailed fact instances
$cons^T(\Pi^T)$ maintains the relationship of entailment; if a fact
instance $f$ is entailed by $\langle R, F^T\rangle$, then $strip(f)$
will be entailed by $\langle R, strip(F^T) \rangle$.
\end{definition}

\begin{definition}[Time-Annotated Justification]
\labeldef{continuous datalog: instance justification}

The justification of entailments (as in \refdef{continuous datalog:
  justification}) is equally applicable to time-annotated Datalog
programs, with facts in the entailment and the justification for that
entailment being replaced with fact instances.

Furthermore, each justification for $f$ in some $F^T$
entails a unique instance ${f | t}$, such that many instances of $f$
may occur in the set $E^T$ of fact instances entailed by ${\Pi^T =
  \langle R, W^{S,x}_i \rangle}$, each with a single justification.

As such, entailment justification for fact instances may be defined
formally by \refdef{continuous datalog: instance justification}.

A justification for some entailment ${f|t}$ of some time-annotated Datalog program
	${\Pi^T = \langle R , F^T \rangle \vdash f|t}$ is the single minimal set ${J^T \subseteq F^T}$
	such that ${\Pi^{T\prime} = \langle R , J^T \rangle \vdash f|t}$.
	The mapping of the fact instance ${f | t}$ to the minimal subset $J^T$
	within ${\Pi^T = \langle R, F^T \rangle}$ that justifies it is expressed
	as the function ${just : \mathbb{\Pi^T} , \mathbb{F^T} \rightarrow \{ \mathbb{F^T} \}}$.
\end{definition}

@@

In \refdef{continuous datalog: datalog program}, we took a Datalog
program to consist of a set of rules and a set of facts.  In order to
extend this definition of Datalog programs to incorporate fact
instances, as in \refdef{continuous datalog: fact instance}, we define
a time-annotated Datalog program as follows:
